Modelling the Lymphatic Metastatic Progression Pathways of OPSCC from Multi-Institutional Datasets

### 2023-01-11 src/tex/ms.tex
[showyourwork] overleaf sync

The transition from one time-step to another is governed by the transition probability $P\left( \mathbf{X}[t+1]=\boldsymbol{\xi}_i \mid \mathbf{X}[t]=\boldsymbol{\xi}_j \right)$, which can conveniently be collected into a transition matrix when we enumerate all $2^V$ distinct possible states $\boldsymbol{\xi}_i$ with $i \in \left\{ 1,2, \ldots, 2^V \right\}$ of lymphatic involvement:
%
\begin{equation}
    \mathbf{A} = \left( A_{ij} \right) = \left( P\left( \mathbf{X}[t+1]=\boldsymbol{\xi}_i \mid \mathbf{X}[t]=\boldsymbol{\xi}_j \right) \right)
\end{equation}
%
The term $P\left( \boldsymbol{\xi}_i \mid \boldsymbol{\xi}_j \right)$ describes the probability to transition from the hidden state of lymphatic involvement $\boldsymbol{\xi}_j$ to the state $\boldsymbol{\xi}_i$ within between the time $t$ and $t+1$. Using a \gls{dag} as depicted in \cref{fig:graph_with_obs}, we can parameterize this transition probability in the following way:
%
\begin{equation}
    \label{eq:transition_prob}
    P\left( \boldsymbol{\xi}_i \mid \boldsymbol{\xi}_j \right) = \prod_{v \leq V} Q\left( \xi_{iv} ; \xi_{jv} \right) P \left( \xi_{iv} \mid \left\{ \xi_{jr} \right\}_{r \in \operatorname{pa}(v)} \right)^{1 - \xi_{jv}}
\end{equation}
%
In this equation, we have denoted \glspl{lnl} that are parents of \gls{lnl} $v$ with the symbol $r\in\operatorname{pa}(v)$. Also, $\xi_{iv}$ denotes the value that \gls{lnl} $v$ takes on when the patient is in state $\boldsymbol{\xi}_i$. Lastly, the term $Q(a;b) \in \{ 0,1 \}$ is there only to

@@

@@

In \cref{fig:IandII_low_prevs,fig:IandII_high_prevs,fig:IIandVII_prevs,fig:IIIandV_prevs,fig:IVandV_prevs} the base graph's predictions for involvement patterns that include the \glspl{lnl} I, V, and VII have been plotted as gray, hatched histograms alongside the corresponding prevalence in the data and the prediction by the winning graph. This allows us to compare how the winning graph improves its fit to the data in these particular scenarios around low-risk \glspl{lnl}. How frequently each of the investigated lymphatic metastatic patterns were observed in the data is additionally tabulated in \cref{table:data_prevalence}, both for early and for late T-category cases.

>>>>

The transition from one time-step to another is governed by the transition probability $P\left( \mathbf{X}[t+1]=\boldsymbol{\xi}_i \mid \mathbf{X}[t]=\boldsymbol{\xi}_j \right)$, which can conveniently be collected into a transition matrix when we enumerate all $2^V$ distinct possible states $\boldsymbol{\xi}_i$ with $i \in \left\{ 1,2, \ldots, 2^V \right\}$ of lymphatic involvement:
%
\begin{equation}
    \mathbf{A} = \left( A_{ij} \right) = \left( P\left( \mathbf{X}[t+1]=\boldsymbol{\xi}_i \mid \mathbf{X}[t]=\boldsymbol{\xi}_j \right) \right)
\end{equation}
%
The term $P\left( \boldsymbol{\xi}_i \mid \boldsymbol{\xi}_j \right)$ describes the probability to transition from the hidden state of lymphatic involvement $\boldsymbol{\xi}_j$ to the state $\boldsymbol{\xi}_i$ between the time $t$ and $t+1$. Using a \gls{dag} as depicted in \cref{fig:graph_with_obs}, we can parameterize this transition probability in the following way:
%
\begin{equation}
    \label{eq:transition_prob}
    P\left( \boldsymbol{\xi}_i \mid \boldsymbol{\xi}_j \right) = \prod_{v \leq V} Q\left( \xi_{iv} ; \xi_{jv} \right) P \left( \xi_{iv} \mid \left\{ \xi_{jr} \right\}_{r \in \operatorname{pa}(v)} \right)^{1 - \xi_{jv}}
\end{equation}
%
In this equation, we have denoted \glspl{lnl} that are parents of \gls{lnl} $v$ with the symbol $r\in\operatorname{pa}(v)$. Also, $\xi_{iv}$ denotes the value that \gls{lnl} $v$ takes on when the patient is in state $\boldsymbol{\xi}_i$. Lastly, the term $Q(a;b) \in \{ 0,1 \}$ is there to prohibit self-healing. It is always one, except if \gls{lnl} $v$ is healthy in state $\xi_i$, but was metastatic in the previous state $\xi_j$. In that case the function becomes $Q(0;1) = 0$, making the transition back to healthier states impossible.

@@

Anatomically, the \emph{posterior accessory pathway} that drains \gls{lnl} II through \gls{lnl} V motivates investigating and arc from $X_2$ into $X_5$ \cite{lengele_anatomical_2007}. And, although no lymphatic pathway is described that directly drains \gls{lnl} III or IV into level V, due to their proximity to each other, we will also investigate additional arcs from \gls{lnl} III and IV into level V.

Lastly, we also compared how the model's performance improves with an added connection from the most commonly involved level II to one of the most rarely metastatic \glspl{lnl} VII.

@@

In \cref{fig:IandII_low_prevs,fig:IandII_high_prevs,fig:IIandVII_prevs,fig:IIIandV_prevs,fig:IVandV_prevs} the base graph's predictions for involvement patterns that include the \glspl{lnl} I, V, and VII have been plotted as gray, hatched histograms alongside the corresponding prevalence in the data and the prediction by the winning graph. Every scenario is show for early and late T-category. This allows us to compare how the winning graph improves its fit to the data in these particular scenarios around low-risk \glspl{lnl}. How frequently each of the investigated lymphatic metastatic patterns were observed in the data is additionally tabulated in \cref{table:data_prevalence}, again both for early and for late T-category cases.
